Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2024-01-17 manuscript_thev_transcriptome.Rmd
added more text to manuscript

---
title: Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  word_document: default
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---

@@

Abraham Quaye^1\*^, Bret Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all vertebrates. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via a highly complex alternative splicing pattern [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs; these are thought to be inherited from a common ancestor. All other genes are termed "genus-specific". "Genus-specific" genes tend to be located at the termini of the genome while "genus-common" genes are usually central [@Davison2003]. This pattern is observed in *Adenoviridae*, *Poxviridae*, and *Herpesviridae* [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) [@Harrach2011; @Guimet2016]. Currently, there are three recognized members of the genus SiAdV: frog adenovirus 1, raptor adenovirus 1, and turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content (\~23 genes) of all known AdVs, and many "genus-specific" putative genes of unknown functions have been annotated (see **Figure 1**)[@Harrach2008; @Kovács2011; @Davison2003].

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV are serologically indistinguishable, infecting turkeys, chickens, and pheasants and the THEV-V cause different clinical diseases in these birds [@Harrach2008; @Beach2006]. In turkeys, the THEV-V cause hemorrhagic enteritis (HE), a debilitating acute disease affecting predominantly 6-12-week-old turkeys characterized by immunosuppression (IS), weight loss, intestinal lesions leading to bloody diarrhea, splenomegaly, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967]. HE is the most economically significant disease caused by any strain of THEV [@Beach2006]. While the current vaccine strain (a THEV-A isolated from a pheasant, Virginia Avirulent Strain [VAS]) have proven effective at preventing HE in young turkey poults, it still retains the immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. The induced IS also interferes with vaccination schemes for other infections of turkeys [@Beach2006; @Rautenschlein2000]. To eliminate this immunossupressive side-effect of the vaccine, a thorough investigation of the culprit viral factors (genes) mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making the investigation of specific viral genes for possible roles in causing IS impractical. A well-characterized transcriptome of THEV is required to enable the next leap forward in THEV research - experimentation with specific viral genes that may mediate IS.

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs - specifically human AdVs - thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP]), which generates five families of late mRNAs (L1-L5). An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. The MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5). Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing of the genes encoded in the unit[@Zhao2014; @Donovan2020; @Guimet2016]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020].

High throughput sequencing methods have facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014]. In this paper, a paired-end deep sequencing experiment was performed to characterize for the first time, the transcriptome of THEV (VAS vaccine strain) during different phases of the infection, yielding the first THEV splicing map. Our paired-end sequencing allowed for reading `r seq_len` bp long high quality (mean Phred Score of 36) sequences from each end of cDNA fragments, which were mapped to the genome of THEV. The generated data from our paired-end sequencing experiment should thus be reliable. \newpage

@@

**Overview of sequencing data and analysis pipeline outputs**\
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel, ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified \#### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (**Figure 3**). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under **accession no. XXXXXX**.

**Changes in THEV splicing profile over time**\
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU) or region. Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs [@Donovan2020; @Zhao2014]. To evaluate the activity of each promoter over time, Firstly, `Ballgown`, a program for statistical analysis of assembled transcriptomes [@Ballgown] was used to estimate and normalize expression levels of all transcripts for each time point in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis.

@@

Individually, TRXPT_21 (DBP) - from the E2 region - was the most significantly expressed at 12 h.p.i, comprising about `r abn_each_12`% of the total transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region; transcripts TRXPT_10 and TRXPT_14 were most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (**Figure 4a**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (**Figure 4b**). Secondly, we estimated relative abundances of all splice junctions for each time point using the raw reads. We counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads counted at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLP was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junctions reads. (**Table 2a**). The levels of the top most abundant junctions at 12 h.p.i were maintained also 24 h.p.i as the most significantly expressed. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b**). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions were not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (**Table 2c**. Also see **Supplementary Tables 1a-c**; **Figure 4c**). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (**Figure 4d**).\
Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

@@

This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 5**). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered. The 5'-most CDS of TRXPT_1 is multi-exonic, encoding a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region - a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but not with Hyd, as the SSC of Hyd is not in-frame. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our RNA-seq data, all E1 transcripts share the same transcription termination site (TTS; at position 2325bp); however, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2 (at position 54bp). Given that studies in MAdVs show that E1 mRNAs share not only a common TTS but also the TSS, and only differ from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and their actual TSS just like the TTS are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential (CP) remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. The CP of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its CP is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (**Figure 5b**; **supplementary PCR methods**). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (**supplementary PCR methods**). This is corroborated by 3'-RACE experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing with a poly-A tail immediately after the TTS of transcripts in this region. The 5'-most CDS of this transcript would encode ORF1. However, the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, suggesting that the predicted ORF1 CDS is a truncated ORF; it shares the same TSS, SSC, and TTS as TRXPT_2, but has a unique STC.

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**     
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and E2B and encodes three classical AdV proteins essential for genome replication: pTP and Ad-pol (E2B proteins), and DBP (E2A protein) [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) were discovered [@Donovan2020], we discovered only a single promoter from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 6**).

>>>>

---
title: Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  word_document: default
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---

@@

Abraham Quaye^1\*^, Brett Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all vertebrates. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via highly complex alternative splicing patterns [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs; these are thought to be inherited from a common ancestor. All other genes are termed "genus-specific". "Genus-specific" genes tend to be located at the termini of the genome while "genus-common" genes are usually central [@Davison2003]. This pattern is observed in *Adenoviridae*, *Poxviridae*, and *Herpesviridae* [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) [@Harrach2011; @Guimet2016]. Currently, there are three recognized members of the genus SiAdV: frog adenovirus 1, raptor adenovirus 1, and turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content (\~23 genes) of all known AdVs, and many "genus-specific" putative genes of unknown functions have been annotated (see **Figure 1**)[@Harrach2008; @Kovács2011; @Davison2003].

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV are serologically indistinguishable, infecting turkeys, chickens, and pheasants, with the THEV-V causing different clinical diseases in these birds [@Harrach2008; @Beach2006]. In turkeys, the THEV-V cause hemorrhagic enteritis (HE), a debilitating acute disease affecting predominantly 6-12-week-old turkeys characterized by immunosuppression (IS), weight loss, intestinal lesions leading to bloody diarrhea, splenomegaly, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967]. HE is the most economically significant disease caused by any strain of THEV [@Beach2006]. While the current vaccine strain (a THEV-A isolated from a pheasant, Virginia Avirulent Strain [VAS]) has proven effective at preventing HE in young turkey poults, it still retains the immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. The induced IS also interferes with vaccination schemes for other infections of turkeys [@Beach2006; @Rautenschlein2000]. To eliminate this immunossupressive side-effect of the vaccine, a thorough investigation of the culprit viral factors (genes) mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making the investigation of specific viral genes for possible roles in causing IS impractical. A well-characterized transcriptome of THEV is required to enable experimentation with specific viral genes that may mediate IS.

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs --- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. Noteworthily, the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but reaches its maximal activity during late infection (after DNA replication). However, during early infection the repertoire of late transcripts from the MLP is restricted until late infection [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. The MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020].

High throughput sequencing methods have facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014]. In this paper, a paired-end deep sequencing experiment was performed to characterize for the first time the transcriptome of THEV (VAS vaccine strain) during different phases of the infection, yielding the first THEV splicing map. Our paired-end sequencing allowed for reading `r seq_len` bp long high quality (mean Phred Score of 36) sequences from each end of cDNA fragments, which were mapped to the genome of THEV.
\newpage

@@

**Overview of sequencing data and analysis pipeline outputs**\
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M. gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**).

Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` novel transcripts. ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified #### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (**Figure 3**).

We validated the splice junctions in all transcripts by PCR amplification of viral cDNA, cloning, and Sanger sequencing (**Supplementary PCR methods**). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under **accession no. XXXXXX**.

**Changes in THEV splicing profile over time**\
AdV gene expression occurs under exquisite temporal control with each promoter typically producing one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts. To evaluate the activity of each promoter over time, `StringTie` and `Ballgown` (a program for statistical analysis of assembled transcriptomes) [@Ballgown] were used to estimate and normalize expression levels of all transcripts for each time point in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis.

@@

Individually, TRXPT_21 (DBP) --- from the E2 region --- was the most significantly expressed at 12 h.p.i, comprising about `r abn_each_12`% of the total transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region --- TRXPT_10 and TRXPT_14 were the most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (**Figure 4a**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (**Figure 4b**). Secondly, we estimated relative abundances of all splice junctions at each time point using the raw reads. We counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLTU was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junction reads (**Table 2a**). The top most abundant junctions at 12 h.p.i remained the most significantly expressed at 24 h.p.i also. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b**). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions were not only the most abundant overall --- accounting for `r round(most_sig_72, 1)`% of all junctions reads --- but also contained the most significantly expressed individual junctions (**Table 2c**. Also see **Supplementary Tables 1a-c**; **Figure 4c**). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (**Figure 4d**).   
We also analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

@@

This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 5**). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the predicted Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered as reported for other AdVs [@Donovan2020; @Zhao2014]. The 5'-most CDS of TRXPT_1 is multi-exonic, encoding a novel 17.9 kilodalton (kDa), 160 residue [amino acids (aa)] protein. From its 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region --- a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211bp). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964bp (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd's STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. Thus, the CDS of TRXPT_2 shares substantial protein sequence similarity with ORF1 but not with Hyd, as the SSC of Hyd is not in-frame. Without its splice site removing the ORF1 STC, TRXPT_2 would encode a longer variant of ORF1, starting from an upstream SSC. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. Our RNA-seq data shows that all E1 transcripts share the same transcription termination site (TTS; at position 2325bp); however, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2 (at position 54bp). Given that studies in MAdVs show that E1 mRNAs share not only a common TTS but also the TSS, and only differ from each other regarding the internal splicing [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and their actual TSS just like the TTS are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential (CP) remains unaffected. Its 5'-most CDS, beginning at 1965bp and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein (ORF4). ORF4 was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. Unlike TRXPT_3, the CP of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its CP is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct, novel, multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. The splice junctions of the transcripts in this region (except the junction for TRXPT_4) were validated by cloning of viral cDNA and Sanger sequencing (**Figure 5**; **Supplementary PCR methods**). During the validation of TRXPT_2, ORF1 was present on the agarose gel (an unspliced band size) and Sanger sequencing results as a bona fide transcript (**Supplementary PCR methods**). This was corroborated by our 3'RACE experiment, which showed a transcript (TRXPT_2B) spanning the entire ORF1 and Hyd ORFs without any splicing, with a poly-A tail immediately after the TTS of transcripts in this region. The 5'-most CDS of this transcript (TRXPT_2B) would encode ORF1. However, TRXPT_2B has an upstream and in-frame SSC to the predicted SSC of ORF1, suggesting that the predicted ORF1 CDS is truncated; it shares the same TSS, SSC, and TTS as TRXPT_2, but has a unique STC.

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**     
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and E2B and encodes three classical AdV proteins essential for genome replication: pTP and Ad-pol (E2B proteins), and DBP (E2A protein) [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) are known [@Donovan2020], we discovered only a single TSS (E2 TSS; 18,751bp) from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 6**).

The E2A ORF, DBP is one of three THEV ORFs predicted to be spliced from two exons. The corresponding transcript (TRXPT_21) found in our data matches this predicted splicing pattern precisely but with a non-coding additional exon at the 5'-end (E2-5'UTR) at position 18,684-18,751 bp. Thus, TRXPT_21 is a three-exon transcript encoding DBP (380 residues, 43.3 kDa) precisely as predicted. This transcript (TRXPT_21) was also corroborated in a 3'-RACE experiment. Additionally, from the 3'-RACE, a splice variant of TRXPT_21 which retains the second intron leading to a 2-exon transcript was found. This transcript (TRXPT_21B), albeit longer due to retaining the second intron and possessing a short 3' UTR, encodes a truncated isoform of DBP because the first SSC utilized by TRXPT_21, is followed shortly by STCs in the retained intron, and does not yield any viable product. Utilizing the SSC 173 bp downstream of TRXPT_21's SSC yields a 346 residue, 39.3 kDa product, which is in-frame of DBP but entirely contained in the second exon. TRXPT_21 and TRXPT_21B share a common TTS but TRXPT_21B as seen in our 3'-RACE data, extends 39 bp into an adenine-thymine (A-T) rich sequence before the poly-A tail sequence occur, suggesting this position (16,934bp) as the bona fide E2A TTS (**Figure 6**).
